# Seamus Stein
### QBS 103
### Presentation 1
### July 25

### APOM Gene information

The APOM gene encodes a protein that is an apolipoprotein. The encoded protein is secreted through the plasma membrane but remains membrane-bound. It functions in the transportation of lipids. 

```{r}
# set working directory to data folder
setwd("/Users/seamus/QBS_103_Project/QBS_103/QBS_103_Data")

# read in gene expression and metadata files 
gene_expression <- read.csv(file = "QBS103_finalProject_geneExpression.csv", row.names = 1)

meta_data <- read.csv(file = "QBS103_finalProject_metadata.csv", row.names = 1)

# subset APOM_expression data. 
APOM_expression <- gene_expression["APOM",]
# apply transform to make the rows into columns 
APOM_expression <- as.data.frame(t(APOM_expression))
head(APOM_expression)
```


``` {r}
# Link metadata to APOM gene expression data
metadata_APOM_expression <- cbind(meta_data, APOM_expression)
head(metadata_APOM_expression)
```


Using ggplot2 generate the following plots for your covariates of choice

a) Histogram for gene expression
```{r}
library(tidyverse)
myTheme <- theme(
  panel.border = element_blank(), panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(), 
  # define axis
  axis.line = element_line(colour = "black", linewidth = rel(1)),
  # plot background
  plot.background = element_rect(fill = "white"),
  panel.background = element_blank(),
  legend.key = element_rect(fill = 'white'), 
  # Move legend
  legend.position = 'right',
  
  # font
  text = element_text(family = 'Georgia', size = 14 ), 
  
  # center titles
  plot.title = element_text(hjust = 0.5)
)
```

```{r}
# define color pallete for APOM expression histogram
colorPalette<- c('#FCDCDC')

# plot histogram of APOM expression 
ggplot(metadata_APOM_expression, aes(x = APOM)) +
  geom_histogram(binwidth = 1, col = 'black', fill = colorPalette ) +
  ggtitle(expression("Histogram of "~italic(APOM)~ " Expression")) +
  labs(x = expression(~italic(APOM)~ " Expression"), y = "Frequency of Patients") +
  myTheme
```

b) Scatter plot for gene expression and continuous covariate age

```{r}
# convert character age to numeric
metadata_APOM_expression$age <- as.numeric(metadata_APOM_expression$age)
head(metadata_APOM_expression)
```
```{r}
# scatter plot for gene expression and age
ggplot(metadata_APOM_expression, aes(x = age, y = APOM)) +
  geom_point(color = "navy") +
  labs(x = "Age (years)", y = expression(~italic(APOM)~ "Expression")) +
  ggtitle(expression(~italic(APOM)~ " Expression by Age")) +
  myTheme
```

c) Boxplot of gene expression separated by both categorical covariates 
Categorical covariates: Sex, icu_status


```{r}
# convert unknown sex to NA
metadata_APOM_expression$sex <- if_else(metadata_APOM_expression$sex == " unknown", NA, metadata_APOM_expression$sex)

```


```{r}
# clean the data by filtering so sex has no NA
APOM_clean <- metadata_APOM_expression %>% 
      filter(!is.na(sex))
```



```{r}
#define color pallete for box plot
colorPalette2 <- c("#db6d00", "#B6DBFF")

# create box plot of gene expression separated by ICU status and sex
ggplot(APOM_clean, aes(x = icu_status, y = APOM, fill = sex)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitterdodge(0.2))+
  #geom_jitter() +
  scale_x_discrete(labels = c("Not Admitted", "Admitted")) +
  labs(x = "ICU Status", y = expression(~italic(APOM)~ " Expression")) +
  scale_fill_manual(values = colorPalette2) +
  ggtitle(expression(~italic(APOM)~ " Exprsession by ICU Status and Sex")) +
  myTheme
```


# Presentation 2
## August 8, 2023


```{r}
# function to create histogram. 
#Takes in the data frame and a singular gene of interest 
createHistogram <- function(df, gene) {
    # create label pieces 
    gene_label <- c(gene)
    labNames <- c('Histogram of', "Expression")
    
    # create dynamic title 
    my_title <- eval(bquote(expression(.(labNames[1]) ~ italic(.(gene_label[1])) ~ .(labNames[2]))))
  
    #create dynamic x axis label
    my_x <- eval(bquote(expression(~ italic(.(gene_label[1])) ~ .(labNames[2]))))
    
    #calculate the median of the gene expression
    med_expression <- median(df[,gene])
    
    # calculate the maximum gene expression
    max_expression <- max(df[,gene])
    #create the histogram
    histogram <-  ggplot(df, aes(x = .data[[gene]])) +
                        geom_histogram(binwidth = max_expression/100, col = 'black', fill = colorPalette ) +
                        labs(x = my_x, y = "Frequency of Patients", title = my_title) +
                        geom_vline(xintercept = med_expression, linetype="solid",
                                   color = "#7A9DFA",
                                   linewidth = rel(1)) +
                        myTheme
    # display histogram
    histogram 
}

```




```{r}
# create scatter plot function. 
#Takes in the data frame, gene, continuous variable, x axis variable and the 
#x-axis variable name for title

createScatterplot <- function(df, gene, conVar, xaxis, x_var_label) {
    # create label pieces
    gene_label <- c(gene)
    labNames <- c(x_var_label,'by', "Expression")
    
    # create dynamic title label
    my_title <- eval(bquote(expression(~italic(.(gene_label[1])) ~ .(labNames[3]) ~ .(labNames[2])
                                       ~ .(labNames[1]))))
    
    # create dynamic y axis title 
    my_y <- eval(bquote(expression(~ italic(.(gene_label[1])) ~ .(labNames[3]))))
    
    #create scatter plot
    scatter_plot <- ggplot(df, aes(x = .data[[conVar]], y = .data[[gene]])) +
                  geom_point(color = "navy") +
                  labs(x = xaxis,  y = my_y, title = my_title) +
                  myTheme
  #display scatter plot
  scatter_plot
}
```

```{r}
# create boxplot function, takes in data frame, gene, x-axis categorical variable, categorical  variable for the color
# 2 tick labels (1 parameter each) label for the x axis categorical variable, label for x fill  
createBoxplot <- function(df, gene, xCat, colorCatVar, tick_label_1, tick_label_2, x_catLabel, x_fill_label) {
  # create label pieces
  gene_label <- c(gene)
  labNames <- c("Expression", "by", x_catLabel, "and", x_fill_label)
  
  # create dynamic title label
  my_title <- eval(bquote(expression(~italic(.(gene_label[1])) ~ .(labNames[1]) ~ .(labNames[2])
                                       ~ .(labNames[3]) ~ .(labNames[4]) ~ .(labNames[5]))))
  # create dynamic y axis title 
  y_axis_title <- eval(bquote(expression(~italic(.(gene_label[1])) ~ .(labNames[1]))))
  
  # create boxplot
  myBoxplot <- ggplot(df, aes(x = .data[[xCat]], y = .data[[gene]], fill = .data[[colorCatVar]])) + 
               geom_boxplot(outlier.shape = NA) +
               geom_point(position = position_jitterdodge(0.2))+
               scale_x_discrete(labels = c(tick_label_1, tick_label_2)) +
               labs(x = x_catLabel, y = y_axis_title,
                    title = my_title) +
               scale_fill_manual(values = colorPalette2,name = "Sex", labels = c("Female", "Male")) +
               myTheme
  #display boxplot
  myBoxplot
}
```



### Subset multiple genes for analysis

```{r}
# subset multiple genes data. 
gene_subset <- gene_expression[c("APOM", "APOA2","LCN2"),]
# apply transform to make the rows into colmnns 
gene_subset <- as.data.frame(t(gene_subset))
head(gene_subset)
```

```{r}
# Link metadata to gene subset data
metadata_gene_expression <- cbind(meta_data, gene_subset)
head(metadata_gene_expression)

```

```{r}
# convert unknown sex to NA
metadata_gene_expression$sex <- if_else(metadata_gene_expression$sex == " unknown", NA, metadata_gene_expression$sex)

```

```{r}
# convert character age to numeric
metadata_gene_expression$age <- as.numeric(metadata_APOM_expression$age)
head(metadata_gene_expression)
```


```{r}
# clean the data by filtering so sex has no NA
gene_clean <- metadata_gene_expression %>% 
      filter(!is.na(sex))

head(gene_clean)
```

```{r}
# This function makes the plots and takes in a vector of 1 continuous variable
# and a vector of 2 categorical variables

makePlots <- function(df, geneNames, x_Cont, xCat) {
  #the x axis categorical variable in box plot
  x_cat <- xCat[1]
  # the categorical variable indicated by color fill
  cat_col = xCat[2]
  
  
  # create the plots by calling functions
  myHist <- createHistogram(df, geneNames)
  myScatter <- createScatterplot(df, geneNames, x_Cont, "Age (years)", "Age")
  myBox <- createBoxplot(df, geneNames, x_cat, cat_col, "Not Admited", "Admitted","ICU Status", "Sex")
  
  #store the plots in a list
  plots <- list(myBox, myHist, myScatter)
  
 # display the plots
 plots
}
```


```{r}
# define variables of interest to pass in the make plots function 
categorical <- c("icu_status", "sex")
continous_vars <- c("age")
geneNames <- c("APOM", "LCN2", "APOA2")
```

```{r}
# loop through the gene names and print all of the plots using makePlots function
for(item in 1:length( geneNames)) {
  gene_plots <- makePlots(gene_clean, geneNames[item], continous_vars, categorical)
  
  print(gene_plots)
}

```


### Presentation 3
### August 22, 2023

#### APOM Plots
```{r}
# define variables of interest to pass in the make plots function 
categorical <- c("icu_status", "sex")
continous_vars <- c("age")
geneNames <- c("APOM")


```
```{r}
# function to create histogram. 
#Takes in the data frame and a singular gene of interest 
createApomHistogram <- function(df, gene) {
    # create label pieces 
    gene_label <- c(gene)
    labNames <- c('Histogram of', "Expression")
    
    # create dynamic title 
    my_title <- eval(bquote(expression(.(labNames[1]) ~ italic(.(gene_label[1])) ~ .(labNames[2]))))
  
    #create dynamic x axis label
    my_x <- eval(bquote(expression(~ italic(.(gene_label[1])) ~ .(labNames[2]))))
    
    #calculate the median of the gene expression
    med_expression <- median(df[,gene])
    
    # calculate the maximum gene expression
    max_expression <- max(df[,gene])
    #create the histogram
    histogram <-  ggplot(df, aes(x = .data[[gene]])) +
                        geom_histogram(binwidth = max_expression/100, col = 'black', fill = colorPalette ) +
                        labs(x = my_x, y = "Frequency of Patients", title = my_title) +
                        geom_vline(xintercept = med_expression, linetype="solid",
                                   color = "#7A9DFA",
                                   linewidth = rel(1)) +
                       annotate("text", x =  med_expression*1.6, y = 0.5*max_expression,
                                 label = paste("Median = ", med_expression), color ='#7A9DFA') + 
                        myTheme
    # display histogram
    histogram 
}
```

```{r}
# function to make APOM histogram, box plot and scatter plot - based on previous plot function 
makeApomPlots <- function(df, geneNames, x_Cont, xCat) {
  #the x axis categorical variable in box plot
  x_cat <- xCat[1]
  # the categorical variable indicated by color fill
  cat_col = xCat[2]
  
  
  # create the plots by calling functions
  myHist <- createApomHistogram(df, geneNames)
  myScatter <- createScatterplot(df, geneNames, x_Cont, "Age (years)", "Age")
  myBox <- createBoxplot(df, geneNames, x_cat, cat_col, "Not Admited", "Admitted","ICU Status", "Sex")
  
  #store the plots in a list
  plots <- list(myBox, myHist, myScatter)
  
 # display the plots
 plots
}
```



```{r}
# loop through the gene names and print all of the plots using makeApomPlots function
for(item in 1:length( geneNames)) {
  gene_plots <- makeApomPlots(gene_clean, geneNames[item], continous_vars, categorical)
  
  print(gene_plots)
}
```


#### Create Table 1 
```{r}
gene_clean[, 'sex'] <- as.factor(gene_clean[, 'sex'])
gene_clean[, 'icu_status'] <-as.factor(gene_clean[, 'icu_status'])
gene_clean[, 'mechanical_ventilation'] <-as.factor(gene_clean[, 'mechanical_ventilation'])
str(gene_clean$sex)
str(gene_clean$icu_status)
levels(gene_clean$sex) <- c("Female", "Male")
levels(gene_clean$icu_status) <- c("Not Admitted", "Admitted")
str(gene_clean$mechanical_ventilation)
levels(gene_clean$mechanical_ventilation) <- c("Not on Mechanical Ventilation", "On Mechanical Ventilation")
```


```{r}
gene_clean$ferritin.ng.ml. <- as.integer(gene_clean$ferritin.ng.ml.)
gene_clean$lactate.mmol.l. <- as.integer(gene_clean$lactate.mmol.l.)

```


```{r}
library(knitr)
library(kableExtra)
# This is function is from the QBS 103 class discussion 
# df = the data frame, my_vars is the list of all variables in the plot
#not notNormContVariables are not normal continuous vars resulting in median and IQR
# cat_var_list are the categorical variable subset where we want n and % values

makeTableOne <- function(df,my_vars,notNormContVar = '',cat_var_list = '') {
  
  # Define an empty table
  table_one <- matrix(nrow = 0,ncol = 2)
  # Keep track of rows to indent
  row_indented <- c()
  
  # Loop through all variables
  for (var in my_vars) {
    
    # Define vector of variable values
    x <- df[,var]
    
    # Identify if not normal distributed continuous variable
    if (var %in% notNormContVar) {
      
      # Calculate individual values
      myMedian <- round(median(x), digits = 2)
      myIQR1 <- round(quantile(x,1/4),digits = 2)
      myIQR2 <- round(quantile(x,3/4),digits = 2)
      # Combine values
      value <- paste0(myMedian,' [',myIQR1,', ',myIQR2,']')
      
      # Define new row
      next_row <- c(paste0('**',var,'** Median [IQR]'),value)
      
      # Add row to data frame
      table_one <- rbind(table_one,next_row)
      
    }
    
    # Identify if categorical
    if (var %in% cat_var_list) {
      
      # Define new row for overall variable
      next_row <- c(paste0('**',var,'** n (%)'),'')
      # Add row to data frame
      table_one <- rbind(table_one,next_row)
      
      # Loop through levels of variable
      for (level in levels(x)) {
        # Calculate n and %
        n <- sum(x == level)
        percent <- round(n/(length(x)) * 100,digits = 2)
        
        # Combine values
        value <- paste0(n,' (',percent,')')
        
        # Define the next row
        next_row <- c(level,value)
        # Add row to data frame
        table_one <- rbind(table_one,next_row)
        # Add index to indented rows
        row_indented <- c(row_indented,nrow(table_one))
        
        
      }
    }
    
    #  treat as normally distributed if not normally distributed or categorical
    if (!(var %in% c(notNormContVar,cat_var_list))) {
      
      # Calculate individual values
      myMean <- round(mean(x),2)
      mySD <- round(sd(x),2)
      # Combine values
      value <- paste0(myMean,' (',mySD,')')
      
      # Define new row
      next_row <- c(paste0('**',var,'** Mean (sd)'),value)
      
      # Add row to data frame
      table_one <- rbind(table_one,next_row)
      
    }
  }
  
  # Define sample size for heading
  sampleSize <- paste("n =",nrow(df))
  
  # Print formatted table
  kable(x = table_one, caption = 'Table 1',
       col.names = c("Variable",sampleSize),
       align = c('l','r'),escape = T,row.names = F) %>%
       add_indent(positions = row_indented) %>%
      kable_classic()
    
  #return(table_one)
}
```

```{r}
my_varList <- c("icu_status", 'sex','mechanical_ventilation', 'age', 'ferritin.ng.ml.','lactate.mmol.l.','APOM', "LCN2", "APOA2")
my_continous_vars <- c( 'APOM', "LCN2", "APOA2")
my_catVars <- c("sex", "icu_status", "mechanical_ventilation")
```


```{r}
# Account for age being NA 
gene_clean <- gene_clean %>% 
      filter(!is.na(age)) %>% 
      filter(!is.na(ferritin.ng.ml.)) %>% 
      filter(!is.na(lactate.mmol.l.))

#summary(gene_clean)
```


```{r}
# produce Table 1 
makeTableOne(gene_clean, my_varList, my_continous_vars, my_catVars)
#write.csv(table_one, file = "table1.csv")
```

#### Generate the Heatmap for 10 genes

```{r}
# subset 10 genes . 
heat_map_genes <- gene_expression[c("APOM", "APOA2","LCN2", "APOA1", "MPO", "BPI","DEFA1", "DEFA4","PRTN3", "CTSG"),]
# apply transform to make the rows into colmnns 
heat_map_genes <- as.data.frame(t(heat_map_genes))
head(heat_map_genes)

```

```{r}
# Link metadata to gene subset data
heat_map_gene_metadata <- cbind(meta_data, heat_map_genes)
head(heat_map_gene_metadata)

```

```{r}
# convert unknown sex to NA
heat_map_gene_metadata$sex <- if_else(heat_map_gene_metadata$sex == " unknown", NA, heat_map_gene_metadata$sex)

```

```{r}
# convert character age to numeric
heat_map_gene_metadata$age <- as.numeric(heat_map_gene_metadata$age)
head(heat_map_gene_metadata)

# Account for age being NA heat clean is the clean data for heatmap
heat_clean <- heat_map_gene_metadata %>% 
      filter(!is.na(age))
```


```{r}
head(heat_clean)
```
```{r}
# create a subset of the genes for the heatmap
my_genes <- c("APOM", "APOA2","LCN2", "APOA1", "MPO", "BPI","DEFA1", "DEFA4","PRTN3", "CTSG")
gene_heatmap_data <-heat_clean[,my_genes]
head(gene_heatmap_data)
```
```{r}
head(heat_clean)
```

```{r}
library(pheatmap)

# define the tracking bars
trackingBars <- data.frame(row.names = rownames(gene_heatmap_data),
                             'ICU Status' = heat_clean$icu_status,
                              'Sex' = heat_clean$sex)


# define the annotation colors
annotation_colors <- list('ICU Status' = c(' no' = 'green', ' yes' = 'pink'),
                          'Sex' = c(' female' =   "blue", ' male' = "red"))

# plot heat map of 10 genes ln transformed (add 0.1 since can't take log of 0).
pheatmap(log(gene_heatmap_data + 0.1),
         clustering_distance_rows  = 'euclidean',
         clustering_distance_cols  = 'euclidean',
         annotation_row  = trackingBars,
         annotation_colors = annotation_colors,
         show_rownames = F,
         main = "Gene Expression Heatmap\nln(Gene Expression)")

```
```{r}

# summary stats of DEFA4 gene based on the heatmap
summary(gene_heatmap_data$DEFA4)
summary(gene_heatmap_data$LCN2)
summary(gene_heatmap_data$DEFA1)
defa1_max <- max(gene_heatmap_data$DEFA1)
log(defa1_max)
```
```{r}
max(gene_heatmap_data$LCN2) - min(gene_heatmap_data$LCN2)
min(gene_heatmap_data$LCN2)
```
```{r}
#calculate DEFA4 expression range from min to max for analysis
max(gene_heatmap_data$DEFA4) - min(gene_heatmap_data$DEFA4)
```
```{r}
#calculate DEFA4 expression maximum for analysis
max(gene_heatmap_data$DEFA4)
```

### Raincloud Plot
In this section, I will use the gene_clean subset for the 3 genes
```{r}
library(ggdist)
```
```{r}
# define a new theme without a legend
noLegendTheme <- theme(
  panel.border = element_blank(), panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(), 
  # define axis
  axis.line = element_line(colour = "black", linewidth = rel(1)),
  # plot background
  plot.background = element_rect(fill = "white"),
  panel.background = element_blank(),
 
  # remove legend
  legend.position = 'none',
  
  # font
  text = element_text(family = 'Georgia', size = 14 ), 
  
  # center titles
  plot.title = element_text(hjust = 0.5)

)
```


```{r}
# A rain cloud plot uses ggplot2 and adds a density plot and a box plot with jitters
# Adapted from https://statsandr.com/blog/graphics-in-r-with-ggplot2/ with the parameter recommendations
APOM_rain_cloud <- gene_clean %>% 
  ggplot(aes(x = icu_status, y = APOM, fill = icu_status)) +
  # add density layer
  stat_halfeye(
    adjust = 0.5,
    # change height of density so they do not interfere
    width = .6,
    
    #move curve to the right
    justification = -0.2,
    
    #remove default interval for cleaner look
    .width = 0,
    point_colour = NA ,
  
  ) + #modify boxplot parameters
  geom_boxplot(
    width = 0.12,
    outlier.color = NA,
    alpha = 0.5
  ) + #modify point parameters
  geom_point(aes(color = icu_status),
            position = position_jitterdodge(0.2),
            size = 1.4,
            alpha = 0.3,
            width = 0.09) +

  # flip the coordinates
  coord_flip() +
  labs(x = "ICU Status", y  = expression(~italic(APOM)~ " Expression"),
       title =  expression(~italic(APOM)~ " Expression by ICU Status")) +
  scale_fill_manual(values = colorPalette2) + 
  #use theme with no legend
  noLegendTheme 


APOM_rain_cloud
```
```{r}
# A rain cloud plot uses ggplot2 and adds a density plot and a box plot with jitters
# function to make a raincloud plot with ICU status as a categorical variable given data frame and gene
make_raincloud <- function(df, gene){
   # create label pieces 
    gene_label <- c(gene)
    labNames <- c("Expression", "by ICU Status")
    
    # create dynamic title 
    my_title <- eval(bquote(expression(italic(.(gene_label[1])) ~ .(labNames[1])~ .(labNames[2]))))
  
    #create dynamic y axis label
    my_y_lab<- eval(bquote(expression(~ italic(.(gene_label[1])) ~ .(labNames[1]))))
  
  
  APOM_rain_cloud <- gene_clean %>% 
  ggplot(aes(x = icu_status, y = .data[[gene]], fill = icu_status)) +
  # add density layer
  stat_halfeye(
    adjust = 0.5,
    # change height of density so they do not interfere
    width = .6,
    
    #move curve to the right
    justification = -0.2,
    
    #remove default interval for cleaner look
    .width = 0,
    point_colour = NA ,
  
  ) + # modify geom_boxplot parameters
  geom_boxplot(
    width = 0.12,
    outlier.color = NA,
    alpha = 0.5
  ) +
  geom_point(aes(color = icu_status),
            position = position_jitterdodge(0.2),
            size = 1.4,
            alpha = 0.3) +
# flip the coordinates
  coord_flip() +
    
  labs(x = "ICU Status", y  = my_y_lab,
       title = my_title) +
  scale_fill_manual(values = colorPalette2) + 
  #use theme with no legend
  noLegendTheme 


APOM_rain_cloud
}
```

```{r}
# create a raincloud plot for each of the 3 genes
the_genes <- c("APOM", "LCN2", "APOA2")
for(item in 1:length(the_genes)) {
  gene_plots <- make_raincloud(gene_clean, the_genes[item])
  
  print(gene_plots)
}

```

#### References
https://stackoverflow.com/questions/31927984/using-italic-with-a-variable-in-ggplot2-title-expression
https://www.appsloveworld.com/r/100/2/combining-paste-and-expression-functions-in-plot-labels
https://statisticsglobe.com/add-mean-and-median-to-histogram-in-r
https://www.biostars.org/p/285370/
https://statsandr.com/blog/graphics-in-r-with-ggplot2/




