# Seamus Stein
### QBS 103
### Presentation 1
### July 25

```{r}
# set working directory to data folder
setwd("/Users/seamus/QBS_103_Project/QBS_103/QBS_103_Data")

# read in gene expression and metadata files 
gene_expression <- read.csv(file = "QBS103_finalProject_geneExpression.csv", row.names = 1)

meta_data <- read.csv(file = "QBS103_finalProject_metadata.csv", row.names = 1)

# subset APOM_expression data. 
APOM_expression <- gene_expression["APOM",]
# apply transform to make the rows into colmnns 
APOM_expression <- as.data.frame(t(APOM_expression))
head(APOM_expression)
```


``` {r}
# Link metadata to APOM gene expression data
metadata_APOM_expression <- cbind(meta_data, APOM_expression)
head(metadata_APOM_expression)
```


Using ggplot2 generate the following plots for your covariates of choice

a) Histogram for gene expression
```{r}
library(tidyverse)
myTheme <- theme(
  panel.border = element_blank(), panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(), 
  # define axis
  axis.line = element_line(colour = "black", linewidth = rel(1)),
  # plot background
  plot.background = element_rect(fill = "white"),
  panel.background = element_blank(),
  legend.key = element_rect(fill = 'white'), 
  # Move legend
  legend.position = 'right',
  
  # font
  text = element_text(family = 'Georgia', size = 14 ) 
)
```

```{r}
# define color pallete for APOM expression histogram
colorPalette<- c('#FCDCDC')

# plot histogram of APOM expression 
ggplot(metadata_APOM_expression, aes(x = APOM)) +
  geom_histogram(binwidth = 1, col = 'black', fill = colorPalette ) +
  ggtitle("Histogram of APOM expression") +
  labs(x = "APOM Expression", y = "Frequency") +
  myTheme
```

b) Scatter plot for gene expression and continuous covariate age

```{r}
# convert character age to numeric
metadata_APOM_expression$age <- as.numeric(metadata_APOM_expression$age)
head(metadata_APOM_expression)
```
```{r}
# scatter plot for gene expression and age
ggplot(metadata_APOM_expression, aes(x = age, y = APOM)) +
  geom_point(color = "navy") +
  labs(x = "Age (years)", y = "APOM Expression") +
  ggtitle("APOM Expression by Age") +
  myTheme
```

c) Boxplot of gene expression separated by both categorical covariates 
Categorical covariates: Sex, icu_status


```{r}
# convert unknown sex to NA
metadata_APOM_expression$sex <- if_else(metadata_APOM_expression$sex == " unknown", NA, metadata_APOM_expression$sex)

```


```{r}
# clean the data by filtering so sex has no NA
APOM_clean <- metadata_APOM_expression %>% 
      filter(!is.na(sex))
```

```{r}
#define color pallete for box plot
colorPalette2 <- c("#db6d00", "#B6DBFF")

# create box plot of gene expression separated by ICU status and sex
ggplot(APOM_clean, aes(x = icu_status, y = APOM, fill = sex)) + 
  geom_boxplot() +
  scale_x_discrete(labels = c("Not Admitted", "Admitted")) +
  labs(x = "ICU Status", y = "APOM expression") +
  scale_fill_manual(values = colorPalette2) +
  ggtitle("APOM Exprsession by ICU Status and Sex") +
  myTheme
```


### APOM Gene information

The APOM gene encodes a protein that is an apolipoprotein. The encoded protein is secreted through the plasma membrane but remains membrane-bound. It functions in the transportation of lipids.   